Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-05-07 draft.tex
Unpushed changed that ended up in the PRD submission

%% Shortcuts %%
\newcommand{\aver}[1]{\langle #1 \rangle}
\newcommand{\dual}[1]{\tilde{#1}}
\newcommand{\est}[1]{\widehat{#1}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\nuvec}{\vec{\nu}}
\newcommand{\order}[1]{\mathcal{O}\left({#1}\right)}
\newcommand{\refapp}[1]{app.~(\ref{app:#1})}
\newcommand{\refeq}[1]{eq.~(\ref{eq:#1})}
\newcommand{\reffig}[1]{fig.~(\ref{fig:#1})}
\newcommand{\refsec}[1]{sec.~(\ref{sec:#1})}
\newcommand{\reftab}[1]{tab.~(\ref{tab:#1})}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\subd}{_{\text{d}}}
\newcommand{\subt}{_{\text{t}}}
\newcommand{\thvec}{\vec{\vartheta}}
\let\oldtheta\theta
\renewcommand{\theta}{\vartheta}
\let\eps\varepsilon
\newcommand{\vecest}[1]{\widehat{\vec{#1}}}
\newcommand{\wwhat}[1]{\widehat{\widehat{#1}}}
\newcommand{\xd}{x\subd}
\newcommand{\xt}{x\subt}
\newcommand{\Ekernel}{E(\xd|\xt)}

@@

Apparently we now face a circular dependence. On the one hand,
the estimators $\vecest{Q}$ and thus also $\vecest{S}$
are proportional to $R$, the ratio of detected events over occuring events.
On the other hand, $R$ depends by construction (see \refeq{def-R})
on $P_\text{t}(x_t)$, and thus on the true value of the angular observables $\vec{S}$.
This dependence is broken by the fact that the MC estimators $\vecest{S}$ need to fulfill
the self-consistency condition
\begin{equation}
    \est{S}_0 = \frac{1}{2}\,,\qquad\text{for a uniangular distribution}.
\end{equation}
(For the process-dependent conditions in the multiangular case see \refeq{recipe:lambdabtolambdall} and
\refeq{recipe:btokstarll}.)
This self-consistency condition is tightly related to the determination of
the branching ratio of the underlying decay, and we therefore suggest to
carry out a combined analysis for the determination of the branching ratio
and the extraction of the angular observables
Moreover, in the applications to $B$ decays only ratios and similar $R$-independent combinations
of the angular observables are of interest.
\\

@@

The accuracy of the unfolding process as outlined above critically depends on
both the accuracy of the detector simulation, as well as the uncertainties
induced by the MC estimates.  For an experimental analysis, one would now turn
to the determination of the distribution of $\vecest{S}$ as a function of the
detector setup. This would involve the determination of both $M$ and
$\vecest{Q}$ for a number of detector configurations, and subsequent profiling
or marginalization.  While such considerations of any detector simulation are
beyond the scope of this work, we can, however, comment on the MC-induced
uncertainties. As usual, one needs to find a balance between compute time and
accuracy. For a uniangular distribution and $\order{10^6}$ MC samples, we find
that the error on the mean of each matrix element is $\order{10^{-4}}$.
This suggests that the so-induced systematic error can be driven below any
statistical uncertainty.

@@

We thank Ulrik Egede for insightful discussions and helpful comments
on the manuscript. We are grateful to Gudrun Hiller, Martin Jung, and
Konstantinos Petridis for helpful discussions, as well as Claus Grupen
for helpful advice in searching the literature.  Tobias Schlüter made
us aware of existing ideas on our method in~\cite{Chung:1997qd} and
his thesis~\cite[page 123]{Schlueter:2012}. The work of D.v.D has been
supported by the Bundesministerium f\"ur Bildung und Forschung
(BMBF). M.C. has been supported by the ''Diamond Grant'' founded by
the Polish Ministry of Higher Education. F.B. is grateful for the
support by the DFG Cluster of Excellence "Origin and Structure of the
Universe". Part of the numerical simulations have been carried out on
the computing facilities of the Computational Center for Particle and
Astrophysics (C2PAP).

>>>>

%% Shortcuts %%
\newcommand{\aver}[1]{\langle #1 \rangle}
\newcommand{\dual}[1]{\tilde{#1}}
\newcommand{\est}[1]{\widehat{#1}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\nuvec}{\vec{\nu}}
\newcommand{\order}[1]{\mathcal{O}\left({#1}\right)}
\newcommand{\refapp}[1]{appendix~\ref{app:#1}}
\newcommand{\refeq}[1]{eq.~(\ref{eq:#1})}
\newcommand{\reffig}[1]{figure~\ref{fig:#1}}
\newcommand{\refsec}[1]{section~\ref{sec:#1}}
\newcommand{\reftab}[1]{table~\ref{tab:#1}}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\subd}{_{\text{d}}}
\newcommand{\subt}{_{\text{t}}}
\newcommand{\thvec}{\vec{\vartheta}}
\let\oldtheta\theta
\renewcommand{\theta}{\vartheta}
\let\eps\varepsilon
\newcommand{\vecest}[1]{\widehat{\vec{#1}}}
\newcommand{\wwhat}[1]{\widehat{\widehat{#1}}}
\newcommand{\xd}{x\subd}
\newcommand{\xt}{x\subt}
\newcommand{\Ekernel}{E(\xd|\xt)}

@@

Apparently we now face a circular dependence. On the one hand,
the estimators $\vecest{Q}$ and thus also $\vecest{S}$
are proportional to $R$, the ratio of detected events over occuring events.
On the other hand, $R$ depends by construction (see \refeq{def-R})
on $P_\text{t}(x_t)$, and thus on the true value of the angular observables $\vec{S}$.
This dependence is broken by the fact that the MC estimators $\vecest{S}$ need to fulfill
the self-consistency condition
\begin{equation}
    \est{S}_0 = \frac{1}{2}\,,\qquad\text{for a uniangular distribution}.
\end{equation}
(For the process-dependent conditions in the multiangular case see \refeq{recipe:lambdabtolambdall} and
\refeq{recipe:btokstarll}.)
This self-consistency condition is tightly related to the determination of
the branching ratio of the underlying decay, and we therefore suggest to
carry out a combined analysis for the determination of the branching ratio
and the extraction of the angular observables.
Moreover, in the applications to $B$ decays only ratios and similar $R$-independent combinations
of the angular observables are of interest.
\\

@@

The accuracy of the unfolding process as outlined above critically depends on
both the accuracy of the detector simulation, as well as the uncertainties
induced by the MC estimates.  For an experimental analysis, one would now turn
to the determination of the distribution of $\vecest{S}$ as a function of the
detector setup. This would involve the determination of both $M$ and
$\vecest{Q}$ for a number of detector configurations, and subsequent profiling
or marginalization.  While such considerations of any detector simulation are
beyond the scope of this work, we can, however, comment on the MC-induced
uncertainties. As usual, one needs to find a balance between compute time and
accuracy. For a uniangular distribution and $\order{10^6}$ MC samples, we find
that the error on the mean of each matrix element is $\order{10^{-4}}$.
This suggests that the so-induced systematic error can be driven below any
statistical uncertainty.\\
An alternative method to unfold detector effects is weighting the data on an
event-by-event basis, with each weight corresponding to the inverse of the detection
efficiency.\\

@@

We thank Ulrik Egede for insightful discussions and helpful comments
on the manuscript. We are grateful to Gudrun Hiller, Martin Jung, and
Konstantinos Petridis for helpful discussions, as well as Claus Grupen
for helpful advice in searching the literature.
Tobias Schlüter made us aware of existing ideas on our method in~\cite{Chung:1997qd} and
his thesis~\cite[page 123]{Schlueter:2012}.
F.B. is grateful for the support by the DFG Cluster of Excellence "Origin and Structure of the
Universe".
M.C. has been supported by the ''Diamond Grant'' founded by
the Polish Ministry of Higher Education.
N.S. acknowledges the support of the Swiss National Science Foundation,
PP00P2-144674.
The work of D.v.D has been supported by the Bundesministerium f\"ur Bildung und Forschung
(BMBF).
Part of the numerical simulations have been carried out on
the computing facilities of the Computational Center for Particle and
Astrophysics (C2PAP).
